Language Bias in a Multilingual Tweet Corpus

### 2018-06-20 paper.tex
Initial review comments

%\usepackage{soul}
%\definecolor{hlcolor}{RGB}{255, 221, 163}
%\sethlcolor{hlcolor}
\newcommand{\hl}[1]{#1}

@@

This paper presents a multilingual, location-anchored corpus of tweets from Latvia. The main challenge of collecting a socially representative corpus from this country is that several languages are used there: Latvian, the main communication language, Russian, the language of the largest minority in Latvia, and English.

%%%%%%%%%
% Problem
%%%%%%%%%
%There are several challenges of building a tweet collection that adequately represents language use in the country.
%
Building a monolingual Latvian collection could be done by harvesting tweets that contain indicative Latvian words which do not present in other languages, similarly to how it is done for Dutch \cite{sang2013}. However, such an approach is not suitable for tweets in Russian and English, as these languages are widely used outside of Latvia.
%
For the same reason, a TREC-like collection building approach \cite{lin2016overview} of filtering the publicly available stream of tweets by language would not work.
%
A tweet collection based on a curated list of users \cite{SANVICENTE16.465} is effective, but extra care must be taken as a thematic bias might be introduced.
%
%A geo-location based collection produces representative results \cite{milajevs:2017:BUCC}.
%
An alternative approach would be to retrieve only geo-located tweets. Such a collection would neither be biased linguistically because it is not based on a list of keywords, nor it would be biased thematically because it is not based on a list of users. The downside is that a large number of tweets are not geo-located, which makes retrieval incomplete.
% IMS self-cite alert. What's new in this work?

@@

The analysis shows that---despite our attempt---the extended collection is biased toward content in Latvian: proportionally there is more tweets in Latvian in the extended collection than in geo-located. We observe this phenomena not only in the ``global level'' (the extended collection), but also at ``micro'' level (across the topics). In all but one topic, Latvian is more dominant than expected.
% IMS I feel like this paragraph is hard to understand because the notion of "topics" hasn't been really defined yet.  Since this is the introduction, the solution might be to make this statement more general.

@@

Over the period from 15 April 2017 to \hl{8 May 2018} the initial set of \hl{1\,708\,236} tweets was collected from the \texttt{POST status/filter} endpoint of the Twitter Streaming API.%\footnotemark{}
%
%\footnotetext{\url{https://developer.twitter.com/en/docs/tweets/filter-realtime/api-reference/post-statuses-filter}}
%
The \texttt{locations} parameter was set to the bounding box of Riga, the capital of Latvia.%\footnotemark{}
%
%\footnotetext{The coordinates of the bounding box are \texttt{23.9325829, 56.8570671, 24.3247299, 57.0859184}.}
% IMS according to http://worldpopulationreview.com/countries/latvia-population/, the population of Riga is less than half the total population of Latvia.  However, Twitter only allows bounding boxes for location to be so big, so you can't bound the whole country.

@@

Russian tweets constitute of more then half tweets coming from TVNET, where its share is \hl{52.8\%} while Latvian is \hl{46.1\%}, and from Facebook, where the two most active users are Russian media.

@@

To investigate the collection further nine topics of interests were defined. The collection was manually searched using keywords to get relevant tweets. The topics are:
\begin{itemize}
\item \texttt{LV001: Winter} Everything about winter.
\item \texttt{LV002: 
The Chronicles of Melanie} Mel\=anijas hronika (The Chronicles of Melanie) is a Latvian movie that was selected for Foreign-Language Category for Oscars.
\item \texttt{LV003: Refugees} A broad topic about refugee crisis in Europe, attitude to immigration and immigrants.
\item \texttt{LV004: Blizzard of Souls} A Latvian movie Dv\=eese\c{l}u putenis (Blizzard of Souls) that is currently in production.
\item \texttt{LV005: Ice hockey} Ice hockey related tweets.
\item \texttt{LV006: Christmas} Christmas.
\item \texttt{LV007: New Year} New Year.
\item \texttt{LV008: Olympics} Tweets about the Olympic games and the Latvian team.
\item \texttt{LV009: Brexit} Brexit.
\end{itemize}

@@

In case of Christmas, we see unexpected activity in March which is due to the discussion of making the Orthodox Christmas a public holiday in Parliament. This topic exhibits some cultural differences and language use. Tweets in Latvian peak during Advent and Christmas in December, tweets in Russian reach maximum both in December and early January when Orthodox Christmas are celebrated. It worth noting that during the discussion whether Orthodox Christmas should be a public holiday, it sometimes was referred as ``Russian Christmas.''

@@

Table~\ref{tab:topic_lang_counts} shows the total number of tweets per topic and language share for Latvian, Russian and English. For all topics the share of Latvian tweets is higher than on average in the collection (Section~\ref{sec:language}). For Russian and English these numbers are generally lower than global average.

>>>>

\usepackage{soul}
\definecolor{hlcolor}{RGB}{255, 221, 163}
\sethlcolor{hlcolor}
%\newcommand{\hl}[1]{#1}

@@

This paper presents a multilingual, location-anchored corpus of tweets from Latvia. The main challenge of collecting a socially representative corpus from this country is that several languages are used there: Latvian, the main communication language and the only official language in Latvia, Russian, the language of the largest minority in Latvia, and English.

%%%%%%%%%
% Problem
%%%%%%%%%
%There are several challenges of building a tweet collection that adequately represents language use in the country.
%
Building a monolingual Latvian collection could be done by harvesting tweets that contain indicative Latvian words, which are not present in other languages, similarly to how it is done for Dutch \cite{sang2013}. However, such an approach is not suitable for tweets in Russian and English, as these languages are widely used outside of Latvia.
%
For the same reason, a TREC-like collection building approach \cite{lin2016overview} of filtering the publicly available stream of tweets by language would not work.
%
A tweet collection based on a curated list of users \cite{SANVICENTE16.465} is effective, but extra care must be taken as a thematic bias might be introduced.
%
%A geo-location based collection produces representative results \cite{milajevs:2017:BUCC}.
%
An alternative approach would be to retrieve only geo-located tweets. Such a collection would neither be biased linguistically because it is not based on a list of keywords, nor it would be biased thematically because it is not based on a list of users. The downside is that a large number of tweets are not geo-located, which makes retrieval incomplete.
% IMS self-cite alert. What's new in this work?

@@

The analysis shows that---despite our attempt---the extended collection is biased toward content in Latvian: proportionally there are more tweets in Latvian in the extended collection than in geo-located. We observe this phenomena not only in the ``global level'' (the extended collection), but also at ``micro'' level (across the topics). In all but one topic, Latvian is more dominant than expected.
% IMS I feel like this paragraph is hard to understand because the notion of "topics" hasn't been really defined yet.  Since this is the introduction, the solution might be to make this statement more general.

@@

Over the period from 15 April 2017 to \hl{8 May 2018} the initial set of \hl{1\,708\,236} tweets was collected from the \texttt{POST status/filter} endpoint of the Twitter Streaming API.%\footnotemark{}
%
%\footnotetext{\url{https://developer.twitter.com/en/docs/tweets/filter-realtime/api-reference/post-statuses-filter}}
%
Following \cite{milajevs:2017:BUCC}, the \texttt{locations} parameter was set to the bounding box of Riga, the capital of Latvia.\footnotemark{}
%
\footnotetext{The coordinates of the bounding box are \texttt{23.9325829, 56.8570671, 24.3247299, 57.0859184}.}
% IMS.2 according to http://worldpopulationreview.com/countries/latvia-population/, the population of Riga is less than half the total population of Latvia.  However, Twitter only allows bounding boxes for location to be so big, so you can't bound the whole country.

@@

More than half tweets that come from TVNET are in Russian. Its share is \hl{52.8\%} while Latvian's is \hl{46.1\%}. Same goes for tweets that come from Facebook, where the two most active users are the Russian edision of media portals (\sn{mixnews\_lv} and \sn{Otkrito})

, where the two most active users are Russian media.

@@

To investigate the collection further, \hl{nine} topics of interests were defined. The collection was manually searched using keywords to get relevant tweets. The topics are:
\begin{itemize}
\item \texttt{LV001: Winter} Everything about winter.
\item \texttt{LV002: 
The Chronicles of Melanie} Mel\=anijas hronika (The Chronicles of Melanie) is a Latvian movie that was selected for Foreign-Language Category for Oscars.
\item \texttt{LV003: Refugees} A broad topic about refugee crisis in Europe, attitude to immigration and immigrants.
\item \texttt{LV004: Blizzard of Souls} A Latvian movie Dv\=ese\c{l}u putenis (Blizzard of Souls) that is currently in production.
\item \texttt{LV005: Ice hockey} Ice hockey related tweets.
\item \texttt{LV006: Christmas} Christmas.
\item \texttt{LV007: New Year} New Year.
\item \texttt{LV008: Olympics} Tweets about the Olympic games and the Latvian team.
\item \texttt{LV009: Brexit} Brexit.
\end{itemize}

@@

In case of Christmas, we see unexpected activity in March which is due to the discussion of making the Orthodox Christmas a public holiday in Parliament. This topic exhibits some cultural differences and language use. Tweets in Latvian peak during Advent and Christmas in December, tweets in Russian reach maximum both in December and early January when Orthodox Christmas are celebrated. It is worth noting that during the discussion whether Orthodox Christmas should be a public holiday, it sometimes was referred as ``Russian Christmas.''

@@

Table~\ref{tab:topic_lang_counts} shows the total number of tweets per topic and language share for Latvian, Russian and English. For all topics the share of Latvian tweets is higher than on average in the collection (Section~\ref{sec:language}). For Russian and English these numbers are generally lower than the global average.
